Reinforcement \change{L}earning for \change{O}ptions\\on \change{T}arget \change{V}olatility \change{F}unds\thanks{Supported by ISCRA from Cineca, \email{iscra@cineca.it}.}

### 2023-02-27 tex/m2ef/dnpp_referee_response.tex

 \item \textit{[Both in Abtract and Introduction, the presentation of the problem to deal with is not well structured. As a matter of fact, first the TVS is briefly introduces, then options and TVO are only hinted, and lastly the problem is shortly described, without all these topics are well linked each other. I suggest the authors to improve this presentation.]}

 ...

 \item \textit{[The Authors address the investigated continuous stochastic control problem through two different approaches: an exact one providing the analytical solution of the problem, and an approximate one based on RL. As for the previous point, even in this one the connections between the two solution approaches are not entirely intelligible. In other terms, it is not entirely clear how the two approaches relate each other. Is the approximate one necessary given that the exact approach provides an analytical solution to the problem? What is the "utility" to use an approximate approach given the applicability of an exact one? I suggest the Authors to clarify this issue.]}

The analytical solution applies only for a specific choice of the model for the dynamics of the underlying and under suitable hypotheses on the payoff. We added equation (52) and a sentence in the first paragraph of section 4 to stress this point.  STEFANO, FORSE HAI ANCHE AGGIUNTO QUALCOSA IN INTRODUZIONE?

@@

 \item \textit{[In Section 4, the Authors state that ``[t]he first [RL] method is a specific algorithm developed by us to fit the problem [...]''. As know, an important feature of any RL method is its convergence to the (unknown) best policy, if any. The Authors do not spend words about this characteristic of the specific algorithm they developed and applied, so I ask them to give some detail about this topic.]}

 ...

 \item \textit{[In Subsection 5.1, the Author claim that an appropriate choice of the learning rate starting value constitutes ``a good choice in terms of [...] avoiding over-fitting''. As known, when training a FFNN, standard methods for avoiding over-fitting are based on error metrics calculated over validation sets or through $K$-fold validation approaches. So, it is not clear as the Authors can guarantee no over-fitting without exploiting such methods or similar. I suggest the Authors to carefully clarify this issue.]}

We added to the second paragraph of section 5.1 a sentence which explains how over-fitting is detected.

### 2023-02-27 tex/m2ef/daluiso_nastasi_pallavicini_polo_rl_for_options_on_tvf.tex
Answers to major remarks 2 and 5
This commit also addresses the "make it clearer" at page 17.

\begin{proposition}
When the underlying securities follow a Black and Scholes \change{QUA IO LASCEREI MODEL (REFEREE PROPONE SCENARIO)} model with deterministic rates, the optimal strategy for a non-decreasing European payoff consists in minimizing the local drift function, independently of the current state $I_t$
\begin{equation}
    \alpha^*(t) \coloneqq \argmin_\alpha \frac{\alpha \cdot \mu(t)}{\|\alpha \cdot \nu(t) \|} \; .
\label{eq:BS_optimal_strategy}
\end{equation}
Analogously, the optimal strategy for a non-increasing European payoff consists in maximizing the local drift function:
\begin{equation}
    \alpha^*(t) \coloneqq \argmax_\alpha \frac{\alpha \cdot \mu(t)}{\norm{\alpha \cdot \nu(t)}} \; .
\label{eq:BS_optimal_strategy_put}
\end{equation}
\end{proposition}
The absence of stochastic elements in Equation~\eqref{eq:BS_optimal_strategy} makes the optimal strategy known \textit{a priori} with no \change{numerical} simulation needed \change{to solve the control problem}; in fact \remove{one can solve the optimization problem once for all $\forall t \in \mathcal{T}$ just looking at the market data $\mu(t)$ and $\nu(t)$ for the securities} \change{, once the risky-assets market data $\mu(t)$ and $\nu(t)$, it is sufficient to apply Equation~\eqref{eq:BS_optimal_strategy} $\forall t \in \mathcal{T}$  to retrieve the deterministic optimal strategy}. Once $\alpha^*$ is known, then one can price the payoff by the following BS formula
\begin{equation}
	V_0^{\text{BS}} = BS(F^{\text{TVS}}(0,T;\alpha^*), K, T, \bar{\sigma}, D(0,T;\zeta)) \; ,
\label{eq:bs_closed_solution}\end{equation}
where $F^{\text{TVS}}(t,T;\alpha)$ is the TVS forward curve defined by
\begin{equation}
	F^{\text{TVS}}(t,T;\alpha) = I_t \exp\left[\int_t^T\left(r(u)-\ell_{\alpha} \left(u\right)\right)du\right] \; ,
\label{eq:tvs_forward}\end{equation}
while $BS(F,K,T,\sigma,D)$ is the standard BS formula for a European option with forward curve $F$, strike $K$, time to maturity $T$, volatility $\sigma$ and discount factor $D$.

@@

If we assume a time-dependent BS dynamics for the risky equities ($\mu_t$, $r_t$ and $\nu_t$ deterministic), then $V = V(t,I)$ and all the derivatives w.r.t. $X$ are zero. Therefore the reduced HJB equation is
\begin{equation}
	\frac{\partial V}{\partial t} +  r(t) I \frac{\partial V}{\partial I}  + \frac{1}{2}\bar{\sigma}^2I^2\frac{\partial^2V}{\partial I^2} + \bar{\sigma}I\max_\alpha \left\{ -\frac{\partial V}{\partial I}   \frac{\alpha \cdot \mu(t)}{\norm{\alpha \cdot \nu(t)}}\right\} = 0 \; ;
\end{equation}
if the payoff is non-decreasing in $I$\change{,} by homogeneity of the \remove{SDE} \change{stochastic differential equation} we get that $V$ is non-decreasing\change{.}\remove{; t}\change{T}hus the solution is given by
\begin{equation}
 \alpha^*(t) = \argmin_\alpha \frac{\alpha \cdot \mu(t)}{\|\alpha \cdot \nu(t)\|} \; ,
	\end{equation}
which is the same result expressed in Equation~\eqref{eq:BS_optimal_strategy}. On the other hand, if the payoff is non-increasing in $I$ then the solution will be the $\argmax$.

Conversely, if we deal with a dynamic model for which the derivative contract $V$ depends on $X$ then the volatility versor term, namely the second one, in the $\max$ operator of Equation~\eqref{eq:HJB_general} is no longer zero and thus one must solve the entire control problem numerically.

In this work we tackle the non-trivial case of a local volatility \remove{(LV)} model for the $S_t$-dynamics, such that $\nu_t = \nu(t,S_t)$. We have chosen the LV model since it is well known in financial literature and among practitioners \change{\cite{Gatheral2006}}.

\section{Reinforcement \change{L}earning}\label{sec:RL}
As we have discussed in the previous sections, one must resort to numerical approaches to solve the stochastic control problem related to the TVS in the case of general payoffs or risky securities dynamics. The standard approach could be to use classical techniques based on backwards recursion \eqref{eq:recursion}-\eqref{eq:recursion2} such as American Monte Carlo \cite{Longstaff2001}. However, their performances degrade exponentially as the dimension $n$ of the problem increases, making prohibitively costly finding the solution to the problem. In our contribution, we adopt a novel technique which is free from the curse of dimensionality and is gaining popularity in many scientific branches for solving stochastic optimal control problems: Reinforcement Learning (RL) \cite{Sutton2018}.

@@

In the \textsc{OpenAI Baselines} implementation of the PPO, the agent policy is parameterized again by a neural network; as for the previous method we have chosen an FFNN. In particular with PPO we deal with a stochastic policy whose functional form is a multivariate diagonal Gaussian distribution where the mean $\mu^\theta(s_k)$ is the output vector of the FFNN and the log-standard deviation $\log\sigma$ is an external parameter
\begin{equation}
	\pi^\theta(s_k) \sim \mathcal{N}(\mu^\theta(s_k),e^{\log\sigma}) \; .
\end{equation}
As one can observe, $\log\sigma$ is state-independent, but it is reduced as the number of the PPO update iterations increases. The idea is that the log-standard deviation will be higher at the beginning of the training phase in order to guarantee a good exploration of the action space while it will be lower at the end to avoid too much noise in the proximity of the optimal policy.

@@

We try to investigate the following hyper-parameters: the FFNN architecture, in particular which feature between the depth and the width of the network is more important, the activation function and the learning rate of the optimizer. Moreover, we compare the performances of two well-known optimizers in Machine Learning literature: Nadam \cite{Dozat2016} and RMSprop \cite{Hiton2012}. Since we deal with a free allocation strategy $\alpha$ that can assume negative values, we have chosen among the wide variety of activation functions \change{\cite{Goodfellow2016}} the tanh and the elu. In this way, we can analyze the performance of a saturating activation function and a non-saturating one. Firstly we have performed a grid search on the learning rate starting value and we have found $10^{-3}$ a good choice in terms of speed of learning and avoiding over-fitting.
 \begin{figure}[h!]
	\centering
	\includegraphics[width=3.4in]{BS_fine_tuning_Nadam_tanh_grey.eps}
	\includegraphics[width=3.4in]{BS_fine_tuning_Nadam_elu_grey.eps}
	\caption{Learning curves of the direct policy algorithm applied to the Black and Scholes \change{scenario}. The solid lines are the loss functions defined in Equation~\eqref{eq:direct_policy_loss} in function of the number of training epochs. Each continuous line represents a different neural network architecture with tanh activation function (top) and elu activation function (bottom). The optimizer adopted is the Nadam with a learning rate of $10^{-3}$. The horizontal grey-dashed line is the conservative option price obtained by maximizing the TVS drift through Equation~\eqref{eq:analytical_solution_free_strategy}}
	\label{fig:result_bs_nadam}
\end{figure}  
 \begin{figure}[h!]
	\centering
	\includegraphics[width=3.4in]{BS_fine_tuning_RMSprop_tanh_grey.eps}
	\includegraphics[width=3.4in]{BS_fine_tuning_RMSprop_elu_grey.eps}
	\caption{Learning curves of the direct policy algorithm applied to the Black and Scholes \change{scenario}. The solid lines are the loss values defined in Equation~\eqref{eq:direct_policy_loss} in function of the number of training epochs. Each continuous line represents a different neural network architecture with tanh activation function (top) and elu activation function (bottom). The optimizer adopted is the RMSprop with a learning rate of $10^{-3}$. The horizontal grey-dashed line is the conservative option price obtained by maximizing the TVS drift through Equation~\eqref{eq:analytical_solution_free_strategy}}
	\label{fig:result_bs_rmsprop}
\end{figure}

>>>>

\begin{proposition}\label{prop:bs}
When the underlying securities follow a Black and Scholes \change{QUA IO LASCEREI MODEL (REFEREE PROPONE SCENARIO)} model with deterministic rates, the optimal strategy for a non-decreasing European payoff consists in minimizing the local drift function, independently of the current state $I_t$
\begin{equation}
    \alpha^*(t) \coloneqq \argmin_\alpha \frac{\alpha \cdot \mu(t)}{\|\alpha \cdot \nu(t) \|} \; .
\label{eq:BS_optimal_strategy}
\end{equation}
Analogously, the optimal strategy for a non-increasing European payoff consists in maximizing the local drift function:
\begin{equation}
    \alpha^*(t) \coloneqq \argmax_\alpha \frac{\alpha \cdot \mu(t)}{\norm{\alpha \cdot \nu(t)}} \; .
\label{eq:BS_optimal_strategy_put}
\end{equation}
\end{proposition}
The absence of stochastic elements in Equation~\eqref{eq:BS_optimal_strategy} makes the optimal strategy known \textit{a priori} with no \change{numerical} simulation needed \change{to solve the control problem}; in fact \remove{one can solve the optimization problem once for all $\forall t \in \mathcal{T}$ just looking at the market data $\mu(t)$ and $\nu(t)$ for the securities} \change{, once the risky-assets market data $\mu(t)$ and $\nu(t)$, it is sufficient to apply Equation~\eqref{eq:BS_optimal_strategy} $\forall t \in \mathcal{T}$  to retrieve the deterministic optimal strategy}. Once $\alpha^*$ is known, then one can price the payoff by the following BS formula
\begin{equation}
	V_0^{\text{BS}} = BS(F^{\text{TVS}}(0,T;\alpha^*), K, T, \bar{\sigma}, D(0,T;\zeta)) \; ,
\label{eq:bs_closed_solution}\end{equation}
where $F^{\text{TVS}}(t,T;\alpha)$ is the TVS forward curve defined by
\begin{equation}
	F^{\text{TVS}}(t,T;\alpha) = I_t \exp\left[\int_t^T\left(r(u)-\ell_{\alpha} \left(u\right)\right)du\right] \; ,
\label{eq:tvs_forward}\end{equation}
while $BS(F,K,T,\sigma,D)$ is the standard BS formula for a European option with forward curve $F$, strike $K$, time to maturity $T$, volatility $\sigma$ and discount factor $D$.

@@

If we assume a time-dependent BS dynamics for the risky equities ($\mu_t$, $r_t$ and $\nu_t$ deterministic), then $V = V(t,I)$ and all the derivatives w.r.t. $X$ are zero. Therefore the reduced HJB equation is
\begin{equation}
	\frac{\partial V}{\partial t} +  r(t) I \frac{\partial V}{\partial I}  + \frac{1}{2}\bar{\sigma}^2I^2\frac{\partial^2V}{\partial I^2} + \bar{\sigma}I\max_\alpha \left\{ -\frac{\partial V}{\partial I}   \frac{\alpha \cdot \mu(t)}{\norm{\alpha \cdot \nu(t)}}\right\} = 0 \; ;
\end{equation}
if the payoff is non-decreasing in $I$\change{,} by homogeneity of the \remove{SDE} \change{stochastic differential equation} we get that $V$ is non-decreasing\change{.}\remove{; t}\change{T}hus the solution is given by
\begin{equation}\label{eq:HJB_solution}
 \alpha^*(t) = \argmin_\alpha \frac{\alpha \cdot \mu(t)}{\|\alpha \cdot \nu(t)\|} \; ,
\end{equation}
which is the same result expressed in Equation~\eqref{eq:BS_optimal_strategy}. On the other hand, if the payoff is non-increasing in $I$ then the solution will be the $\argmax$.

Conversely, if we deal with a dynamic model for which the derivative contract $V$ depends on $X$ then the volatility versor term, namely the second one, in the $\max$ operator of Equation~\eqref{eq:HJB_general} is no longer zero and thus one must solve the entire control problem numerically. \changeB{Indeed, one can check by direct substitution that the natural generalization of \eqref{eq:HJB_solution}
\begin{equation}
 \alpha^*(X_t) = \argmin_\alpha \frac{\alpha \cdot \mu(X_t)}{\|\alpha \cdot \nu(X_t)\|}
\end{equation}
fails to satisfy \eqref{eq:HJB_general}.}

In \removeB{this work}\changeB{what follows} we tackle the non-trivial case of a local volatility \remove{(LV)} model for the $S_t$-dynamics, such that $\nu_t = \nu(t,S_t)$. We have chosen the LV model since it is well known in financial literature and among practitioners \change{\cite{Gatheral2006}}.

\section{Reinforcement \change{L}earning}\label{sec:RL}
As we have discussed in the previous sections, \changeB{the analytical solution described in Proposition \ref{prop:bs} crucially relies on the assumption that the underlyings evolve according to a Black and Scholes model, and on a monotonicity assumption on the payoff. Therefore,}
one must resort to numerical approaches to solve the stochastic control problem related to the TVS in the case of general payoffs or risky securities dynamics. The standard approach could be to use classical techniques based on backwards recursion \eqref{eq:recursion}-\eqref{eq:recursion2} such as American Monte Carlo \cite{Longstaff2001}. However, their performances degrade exponentially as the dimension $n$ of the problem increases, making prohibitively costly finding the solution to the problem. In our contribution, we adopt a novel technique which is free from the curse of dimensionality and is gaining popularity in many scientific branches for solving stochastic optimal control problems: Reinforcement Learning (RL) \cite{Sutton2018}.

@@

In the \textsc{OpenAI Baselines} implementation of the PPO, the agent policy is parameterized again by a neural network; as for the previous method we have chosen an FFNN. \removeB{In particular with PPO we deal with a stochastic policy, whose functional form is} \changeB{More precisely, training is based on a stochastic policy, i.e. the action is drawn from a probability law, and the neural network is used to determine its parameters. In settings like ours where the action has a continuous domain, such law is} a multivariate diagonal Gaussian distribution where the mean $\mu^\theta(s_k)$ is the output vector of the FFNN and the log-standard deviation $\log\sigma$ is an \removeB{external} \changeB{additional trainable} parameter:
\begin{equation}
	\pi^\theta(s_k) \sim \mathcal{N}(\mu^\theta(s_k),e^{\log\sigma}) \; .
\end{equation}
As one can observe, $\log\sigma$ is state-independent, but it is reduced as the number of the PPO update iterations increases. The idea is that the log-standard deviation will be higher at the beginning of the training phase in order to guarantee a good exploration of the action space while it will be lower at the end to avoid too much noise in the proximity of the optimal policy. \changeB{Eventually at the end of training, the standard deviation is put to zero, so that the action becomes deterministically equal to the output of the FFNN.}

@@

We try to investigate the following hyper-parameters: the FFNN architecture, in particular which feature between the depth and the width of the network is more important, the activation function and the learning rate of the optimizer. Moreover, we compare the performances of two well-known optimizers in Machine Learning literature: Nadam \cite{Dozat2016} and RMSprop \cite{Hiton2012}. Since we deal with a free allocation strategy $\alpha$ that can assume negative values, we have chosen among the wide variety of activation functions \change{\cite{Goodfellow2016}} the tanh and the elu. In this way, we can analyze the performance of a saturating activation function and a non-saturating one. Firstly we have performed a grid search on the learning rate starting value and we have found $10^{-3}$ a good choice in terms of speed of learning and avoiding over-fitting. \changeB{Note that since our environment is simulated, over-fitting is easily detected by cross-validation on a set of freshly-generated scenarios which were not seen during optimization.}
 \begin{figure}[h!]
	\centering
	\includegraphics[width=3.4in]{BS_fine_tuning_Nadam_tanh_grey.eps}
	\includegraphics[width=3.4in]{BS_fine_tuning_Nadam_elu_grey.eps}
	\caption{Learning curves of the direct policy algorithm applied to the Black and Scholes \change{scenario}. The solid lines are the loss functions defined in Equation~\eqref{eq:direct_policy_loss} in function of the number of training epochs. Each continuous line represents a different neural network architecture with tanh activation function (top) and elu activation function (bottom). The optimizer adopted is the Nadam with a learning rate of $10^{-3}$. The horizontal grey-dashed line is the conservative option price obtained by maximizing the TVS drift through Equation~\eqref{eq:analytical_solution_free_strategy}}
	\label{fig:result_bs_nadam}
\end{figure}  
 \begin{figure}[h!]
	\centering
	\includegraphics[width=3.4in]{BS_fine_tuning_RMSprop_tanh_grey.eps}
	\includegraphics[width=3.4in]{BS_fine_tuning_RMSprop_elu_grey.eps}
	\caption{Learning curves of the direct policy algorithm applied to the Black and Scholes \change{scenario}. The solid lines are the loss values defined in Equation~\eqref{eq:direct_policy_loss} in function of the number of training epochs. Each continuous line represents a different neural network architecture with tanh activation function (top) and elu activation function (bottom). The optimizer adopted is the RMSprop with a learning rate of $10^{-3}$. The horizontal grey-dashed line is the conservative option price obtained by maximizing the TVS drift through Equation~\eqref{eq:analytical_solution_free_strategy}}
	\label{fig:result_bs_rmsprop}
\end{figure}
